---
title: "Course Schedule Processing — Tier 0 · Explore"
subtitle: "Detroit Public Schools · config-free data profiling"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(gt)
library(htmltools)
library(Microsoft365R)             # get_sharepoint_site()
library(erstools)                 # ers_read_sharepoint()
source("explore_helpers.R")

# NO config, NO column_map, NO D_ names. Reads the files exactly as received.

orgfiles_site_url <- "https://erstrategies1.sharepoint.com/sites/orgfiles"
orgfiles_site  <- get_sharepoint_site(site_url = orgfiles_site_url)
internal_drive <- orgfiles_site$get_drive("Internal")

raw_data_folder_path <- "District Partners/Detroit Public Schools/26-27 HS Redesign Implementation/1. Data & Analysis - Secure/Fall 2026 Course Schedule Analysis"

# Course schedule — one row per student per course
cs_raw <- ers_read_sharepoint(
  folder_path = raw_data_folder_path,
  file_name_with_extension = "/0. Raw Data/course_sections_251103.csv")

# Location coding — the CS file has school IDs but not school names; these get
# joined in. Profile it here to find the join key and what it provides.
location_coded <- ers_read_sharepoint(
  folder_path = raw_data_folder_path,
  file_name_with_extension = "/1. Coding/location_coded.xlsx")

# Student demographics: NOT received yet — will be profiled the same way when it
# arrives (drives check 0.6 in the review artifact).
```

## What this pass is for

This is the **exploratory** half of Tier 0 — raw files, district headers, no
config. It gives you what you need to write `column_map` in
`district_config.yml` and to set up the school-name join.

**Flow:** raw data → *this* (Explore) → author `column_map` + location join →
Review.

As you read, you're hunting for:

- Which column encodes **time** — a `Period` *and* a `Rotation`/day, or a single
  expression-like column (e.g. `2-8(A)`)? (Answers *expression vs. period+rotation*.)
- The **term** values — FY, semesters, quarters, a mix? (Answers *term structure*.)
- The **student id / teacher id** columns.
- The **school-id column** — and its match in `location_coded`, since that's the
  key you'll join on to bring school **names** into the CS data.
- Anything that looks off (a column that isn't what its name suggests).

## Course schedule — column profile

```{r cs-profile}
explore_gt(explore_profile(cs_raw), "Every column: type · distinct · missing · samples")
```

## Course schedule — value sets (low-cardinality columns)

```{r cs-lowcard}
explore_render(explore_low_card(cs_raw, max_distinct = 25))
```

## Course schedule — peek at candidate time columns

List the columns that look like they encode **time**, then inspect their top
values — this is where an expression hiding in a day/rotation column shows up.

```{r cs-peek}
# EDIT to your headers (term / rotation / period / expression):
peek_cols <- c("Term", "Rotation", "Period", "Expression")

peek <- peek_cols[peek_cols %in% names(cs_raw)]
if (length(peek) == 0) {
  htmltools::HTML("<em>None of the named columns are present — adjust peek_cols to match the profile above.</em>")
} else {
  explore_render(setNames(lapply(peek, function(c) explore_value_counts(cs_raw, c, top = 40)), peek))
}
```

## Location coding — column profile

The join source for school names (and likely school level/type). Find the
column here that matches the school-id column in the CS profile above — that's
your join key.

```{r loc-profile}
explore_gt(explore_profile(location_coded), "Every column: type · distinct · missing · samples")
```

## Location coding — value sets (low-cardinality columns)

```{r loc-lowcard}
explore_render(explore_low_card(location_coded, max_distinct = 40))
```

## From here → column_map, the school join, and (later) demographics

1. **`column_map`** — fill `district_config.yml` from the CS profile: keys are
   ERS `D_` names, values are Detroit's raw headers. Note `D_location_name`
   will *not* be here — it comes from the join, not a rename.
2. **School-name join** — join `location_coded` onto the CS data on the shared
   school-id key so `D_location_name` (and level/type) come in. This is a real
   Tier 0 element: "school coding joined, every school id matched."
3. **Demographics** — when the file arrives, profile it here too; that unlocks
   check 0.6 (SWD/ELL/FRL rates + demo-vs-CS reconciliation) with no rework
   elsewhere.

The mapping skill can draft the `column_map` block and flag the join key + any
missing critical fields straight from these profiles.
