---
title: "Course Schedule Processing — Tier 1 · Explore"
subtitle: "Detroit Public Schools · config-free data profiling"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
# Remove everything from any previous sessions
rm(list = ls())
# Load necessary libraries
library(tidyverse)
library(ggplot2)
#load data
library(openxlsx) # Provides functions for reading, writing, and manipulating Excel files: read.xlsx, write.xlsx
library(here) # A simpler way to manage file paths: here() generates paths relative to the project root
library(Microsoft365R) # Use Microsoft365R to connect to SharePoint and load files
library(readxl) # Provides functions for reading Excel files: read_excel()
#manipulate data
library(tidyverse) # Includes ggplot2, dplyr, tidyr, readr, etc.
library(data.table) # fread(): load large CSVs quickly
library(dplyr) # Functions for data manipulation: filter(), select(), mutate(), etc. (part of tidyverse)
library(safejoin) # Provides safer join operations for 1:many matching - safe_left_join()
library(lubridate) # Provides functions for working with date-times: ymd(), hms(), today()
library(scales) # Provides functions for percents
#format and display data
library(flextable) # Creates and formats tables in a publication-ready format: flextable()
library(grid)
library(gtable)
library(gridExtra)
library(glue)
library(gt)
library(dplyr)
library(stringr)
library(htmltools)
library(Microsoft365R)            # required: get_sharepoint_site() + drive objects
library(erstools)                 # ers_read_sharepoint()
source("00_explore_helpers.R")

# NO config, NO column_map, NO D_ names. Reads the files exactly as received.

orgfiles_site_url <- "https://erstrategies1.sharepoint.com/sites/orgfiles"
orgfiles_site  <- get_sharepoint_site(site_url = orgfiles_site_url)
internal_drive <- orgfiles_site$get_drive("Internal")

raw_data_folder_path <- "District Partners/Detroit Public Schools/26-27 HS Redesign Implementation/1. Data & Analysis - Secure/Fall 2026 Course Schedule Analysis"

# Course schedule — one row per student per course
cs_raw <- ers_read_sharepoint(
  folder_path = raw_data_folder_path,
  file_name_with_extension = "/0. Raw Data/course_sections_251103.csv")

# Location coding — the CS file has school IDs but not school names; join source.
location_coded <- ers_read_sharepoint(
  folder_path = raw_data_folder_path,
  file_name_with_extension = "/1. Coding/location_coded.xlsx")

# Student demographics: NOT received yet — profile it here when it arrives.
```

## What this pass is for

The **exploratory** half of Tier 0 — raw files, district headers, no config.
Two jobs: (1) see what you have well enough to author `column_map`, and (2)
work out the **prep recipe** (join / filter / derive) that shapes the raw file
into the fields the pipeline expects. The loop is: **profile raw → prep edits →
re-profile → iterate** until the prepped profile looks right.

## Course schedule — column profile (raw)

Each column with type, distinct count, missingness, min/max, and — for
low-distinct columns — the full value set inline.

```{r cs-profile}
explore_gt(explore_profile(cs_raw), "Raw course schedule — every column")
```

## Course schedule — peek at candidate time columns

High-cardinality columns show only a sample above. List the ones that look like
they encode **time** (term / rotation / period / expression / course number)
and inspect their top values — including whether term is hiding inside the
course number.

```{r cs-peek}
# EDIT to your headers:
peek_cols <- c("Term", "Rotation", "Period", "Expression", "Course Number", "Period Expression")

peek <- peek_cols[peek_cols %in% names(cs_raw)]
if (length(peek) == 0) {
  htmltools::HTML("<em>None of the named columns are present — adjust peek_cols to match the profile above.</em>")
} else {
  explore_render(setNames(lapply(peek, function(c) explore_value_counts(cs_raw, c, top = 40)), peek))
}
```

## Location coding — column profile (raw)

The join source for school names (and likely level/type). Find the column here
that matches the school-id column in the CS profile — that's your join key.

```{r loc-profile}
explore_gt(explore_profile(location_coded), "Raw location coding — every column")
```

## Prep edits (staging)

Your scratch space to shape the raw data into what the pipeline needs. Each edit
below eventually **promotes** to a durable home — this block is where you get it
right first. Fill in the `<...>` placeholders from the profiles above, then set
this chunk's `eval` to `true`.

```{r prep}
#| eval: true
snapshot_s1 <- as.Date("2025-10-30")
snapshot_s2 <- as.Date("2026-02-15")

cs_prepped <- cs_raw |>
  # (1) RENAME — map district headers to pipeline names.
  rename(
    D_stu_id            = MaskedStudentID,
    D_employee_id       = MaskedTeacherID,
    D_location_id       = EEM_CODE,
    D_expression        = `Period Expression`,
    D_course_name       = `Course Name`,
    D_course_id         = `Course Number`,
    D_sced_subject_code = `SCED Code`,
    D_year_id           = Year_End,
    D_stu_enter_date    = `Date Enrolled`,
    D_stu_exit_date     = `Date Left`
  ) |>

  # (2) JOIN — bring school names in from location_coded.
  filter(!is.na(D_location_id), D_location_id != "NULL") |>
  mutate(D_location_id = as.integer(D_location_id)) |>
  left_join(
    location_coded |>
      mutate(D_location_id = as.integer(D_location_id)),
    by = "D_location_id"
  ) |>

  # (3) FILTER — based on what you find in the initial explore
  filter(C_location_type == "Neighborhood School") |>
  filter(D_expression != "NULL") |>
  filter(D_year_id == 2026) |>

  # (4) DERIVE — term from course number/name.
  mutate(
    # Create term from course-number suffix
    D_course_id_suffix = str_extract(
    str_to_upper(str_trim(D_course_id)),
    "S[12]$"
    ),
    D_term = case_when(
      # First priority: course-name ending
      str_detect(
        str_to_upper(D_course_name),
        "-\\s*A\\s*\\*?\\s*$"
      ) ~ "S1",
      str_detect(
        str_to_upper(D_course_name),
        "-\\s*B\\s*\\*?\\s*$"
      ) ~ "S2",
      # Second priority: course-ID ending
      D_course_id_suffix == "S1" ~ "S1",
      D_course_id_suffix == "S2" ~ "S2",
      # Otherwise, full year
      TRUE ~ "FY"
    ),
    # Student must be enrolled across the applicable term window
    C_stu_snapshot = case_when(
      D_term == "S1" &
        D_stu_enter_date <= snapshot_s1 &
        (is.na(D_stu_exit_date) | D_stu_exit_date >= snapshot_s1) ~ "Include",

      D_term == "S2" &
        D_stu_enter_date <= snapshot_s2 &
        (is.na(D_stu_exit_date) | D_stu_exit_date >= snapshot_s2) ~ "Include",

      D_term == "FY" & (
        D_stu_enter_date <= snapshot_s1 &
          (is.na(D_stu_exit_date) | D_stu_exit_date >= snapshot_s1) |
        D_stu_enter_date <= snapshot_s2 &
          (is.na(D_stu_exit_date) | D_stu_exit_date >= snapshot_s2)
      ) ~ "Include",

      TRUE ~ "Exclude"
    )
  ) |>
  # (5) SELECT - columns that start with eith "D_" or "C_"
  select(starts_with("D_"), starts_with("C_"))
```

## Reinvestigate — profile after edits

Re-profile the prepped data to confirm each edit did what you expected: school
names populated, only the right schools remain, and `D_term` holds valid term
values. Iterate the prep block until this looks right.

```{r prep-profile}
if (exists("cs_prepped")) {
  explore_gt(explore_profile(cs_prepped), "After prep edits — re-profiled")
} else {
  htmltools::HTML("<em>Fill in the prep chunk above and set its eval to true, then re-render.</em>")
}
```

```{r prep-peek}
# EDIT to your headers:
peek_cols <- c("D_term", "D_rotation", "D_period", "D_expression", "D_stu_enter_date", "D_stu_exit_date")

peek <- peek_cols[peek_cols %in% names(cs_prepped)]
if (length(peek) == 0) {
  htmltools::HTML("<em>None of the named columns are present — adjust peek_cols to match the profile above.</em>")
} else {
  explore_render(setNames(lapply(peek, function(c) explore_value_counts(cs_prepped, c, top = 40)), peek))
}
```

## From here → promote the recipe

Once the prepped profile looks right, promote each piece to its durable home so
the pipeline is reusable, not one-off:

- **Renames** → `column_map` in `district_config.yml`.
- **Join** → a config location-join spec (source file · key · columns).
- **Filter** → the Tier 1 schools-in-scope decision (config).
- **Derivations** (like `D_term` from the course number) → a derivations block —
  the thing `column_map` can't express, because these fields are computed.
- **Demographics** → profile it here when it arrives; unlocks review check 0.6.

```{r export}
#| eval: true
# Save course schedule data for next script
ers_write_sharepoint(
  data = cs_prepped,
  folder_path = raw_data_folder_path,
  file_name_with_extension = 
    "/2. Processed Data/01_course_data_post_explore.csv")

```

```{r export coding}
#| eval: false
# Export the coding table (hand-code -> course_coding_2026.xlsx)
cs_coding_data <- cs_prepped |>
  group_by(D_course_name, D_course_id, D_sced_subject_code) |>
  count() |> ungroup() |>
  mutate(C_course_time = "UNCODED", C_course_subject_area = "UNCODED",
         C_course_subject = "UNCODED", C_course_credit_type = "UNCODED",
         C_course_name = "UNCODED", C_course_rigor = "UNCODED",
         C_course_rigor_detail = "UNCODED", C_course_intervention = "UNCODED",
         C_course_time_of_day = "UNCODED", C_course_format = "UNCODED",
         C_course_pathway = "UNCODED")

ers_write_sharepoint(
  data = cs_coding_data,
  folder_path = raw_data_folder_path,
  file_name_with_extension = "/1. Coding/course_coding_data.csv")

```