---
title: "Course Schedule Processing — 02 · Coding & Exclusions Review"
subtitle: "Springfield Public Schools · 2025-26"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
rm(list = ls())
library(gt)
library(dplyr)
library(tidyr)
library(glue)
library(yaml)
library(Microsoft365R)
library(erstools)
source("00_review_helpers.R")
source("00_expression_helpers.R")
source("00_db_helpers.R")
config <- yaml::read_yaml("00_district_config.yml")
raw_data_folder_path <- config$raw_data_folder_path

orgfiles_site_url <- "https://erstrategies1.sharepoint.com/sites/orgfiles"
orgfiles_site  <- get_sharepoint_site(site_url = orgfiles_site_url)
internal_drive <- orgfiles_site$get_drive("Internal")

# Stage 01 output (prepped, D_-named, demographics already merged in 01).
cs_data <- ers_read_sharepoint(
  folder_path = raw_data_folder_path,
  file_name_with_extension = "/2. Processed Data/01_course_data_post_explore.csv")

# Hand-coded course coding. Wrapped so the doc renders before it exists.
cs_course_coding <- tr_safe(
  ers_read_sharepoint(
    folder_path = raw_data_folder_path,
    file_name_with_extension = config$raw_files$course_coding),
  otherwise = NULL)

# --- Config-driven exclusion interpreter (no per-district code edits) ---------
# A record is "Exclude" for a metric if ANY of that metric's rules matches.
# A rule whose column isn't present yet is skipped — never jams.
apply_exclusions <- function(df, rules) {
  for (field in names(rules)) {
    excl <- rep(FALSE, nrow(df))
    for (rule in rules[[field]]) {
      if (!rule$col %in% names(df)) next
      x <- df[[rule$col]]; v <- rule$val
      hit <- switch(rule$op,
                    eq = x == v, ne = x != v,
                    `in` = x %in% v, not_in = !(x %in% v),
                    rep(FALSE, length(x)))
      hit[is.na(hit)] <- FALSE
      excl <- excl | hit
    }
    df[[field]] <- ifelse(excl, "Exclude", "Include")
  }
  df
}
# --- what-if sizing filters (drop rows from DB sizing ONLY; default none) ------
.sizing_match <- function(df, rule) {
  if (!rule$col %in% names(df)) return(rep(FALSE, nrow(df)))
  x <- df[[rule$col]]; v <- rule$val
  hit <- switch(rule$op,
                equals = x == v, not_equals = x != v,
                `in` = x %in% v, not_in = !(x %in% v),
                rep(FALSE, length(x)))
  hit[is.na(hit)] <- FALSE
  hit
}
apply_sizing_filters <- function(df, filters) {
  if (is.null(filters) || length(filters) == 0) return(df)
  drop <- rep(FALSE, nrow(df))
  for (rule in filters) drop <- drop | .sizing_match(df, rule)
  df[!drop, , drop = FALSE]
}
sizing_filter_summary <- function(df, filters) {
  if (is.null(filters) || length(filters) == 0)
    return(data.frame(rule = "(none — no sizing filters set)", rows_removed = 0L))
  do.call(rbind, lapply(filters, function(rule)
    data.frame(rule = paste(rule$col, rule$op, paste(rule$val, collapse = " | ")),
               rows_removed = sum(.sizing_match(df, rule)))))
}
# --- Join coding + capture merge diagnostics + derive exclusions --------------
join_key   <- if (!is.null(config$coding$join_key)) config$coding$join_key else "D_course_name"
merge_diag <- NULL
if (!is.null(cs_course_coding)) {
  cs_data <- tr_safe({
    coded_cols <- setdiff(names(cs_course_coding), join_key)
    cc <- cs_course_coding; cc$.coded <- TRUE
    n_before   <- nrow(cs_data)
    key_unique <- !any(duplicated(cc[[join_key]]))
    joined <- cs_data |>
      select(-any_of(coded_cols)) |>
      left_join(cc, by = join_key)
    unmatched <- joined |> filter(is.na(.coded))
    merge_diag <<- list(
      key = join_key, key_unique = key_unique,
      n_before = n_before, n_after = nrow(joined),
      row_stable = identical(n_before, nrow(joined)),
      n_unmatched = nrow(unmatched),
      unmatched_courses = unmatched |> distinct(D_course_name) |> arrange(D_course_name))
    joined |> select(-.coded) |> apply_exclusions(config$exclusions)
  }, otherwise = cs_data)
}

has_cols <- function(...) all(c(...) %in% names(cs_data))

# --- Double-booking sizing (needs coding + exclusions above) ------------------
spec         <- tr_safe(expr_spec(config), NULL)
term_expands <- tr_safe(term_expands_spec(config), NULL)
cs_data$C_class_id <- build_class_id(cs_data, config$class_id$recipe)
cs_size <- apply_sizing_filters(as.data.frame(cs_data), config$db_sizing_filters)

# explode once; flag per the ERS method — students on time-included records,
# teachers on teacher-load-included records.
cs_long <- tr_safe(explode_conflicts(cs_size, spec, term_expands), NULL)
flagged_students <- tr_safe(
  unique(db_flagged(filter(cs_long, C_course_time_exclude == "Include"),
                    "D_stu_id", "C_class_id")$D_stu_id), character(0))
flagged_teachers <- tr_safe(
  unique(db_flagged(filter(cs_long, C_class_size_and_teacher_load_exclude == "Include"),
                    "D_employee_id", "C_class_id")$D_employee_id), character(0))

pct_of <- function(a, b) { r <- round(100 * a / b, 1); r[is.na(b) | b == 0] <- NA_real_; r }
n_stu_time <- tr_safe(n_distinct(cs_data$D_stu_id[cs_data$C_course_time_exclude == "Include"]), NA)
n_tch_load <- tr_safe(n_distinct(cs_data$D_employee_id[cs_data$C_class_size_and_teacher_load_exclude == "Include"]), NA)
```

## How to read this

The first formal review — `01` is investigation-only. Every check either
**computes** a status from the data or is **evidence-only** (a table to eyeball,
no badge). Nothing here is hand-maintained prose wearing a status. Checks detect
what's present and degrade gracefully: a not-yet-loaded input shows "data not
loaded," never a crash.

Golden rule: **processing flags, it never deletes** — exclusions are columns a
later metric can honor or ignore.

## 02 — Coding & exclusions review

```{r stage02}
#| results: 'asis'
s02 <- tr_stage(
  number = "02",
  title  = "Coding & Exclusions Review",
  question = "Did standardization land, do we know what each course is, and is the right thing excluded?",
  checks = list(

    tr_check("2.0", "Did standardization from 01 land — every mapped field present?",
      good   = "Every field in column_map is present with its D_ name.",
      value  = tr_safe(length(setdiff(names(config$column_map$course_schedule), names(cs_data))), NA),
      status = tr_safe(if (length(setdiff(names(config$column_map$course_schedule), names(cs_data))) == 0) "ok" else "flag", "none"),
      evidence = function() {
        f <- names(config$column_map$course_schedule)
        data.frame(field = f, present = ifelse(f %in% names(cs_data), "yes", "MISSING"))
      }),

    tr_check("2.1", "Term structure — is every term value weightable?",
      good   = "Every term in the data has a configured weight.",
      status = tr_safe(if (all(as.character(unique(na.omit(cs_data$D_term))) %in% names(config$weights$term))) "ok" else "flag", "none"),
      evidence = function() {
        cs_data |> count(D_term, name = "n_records") |> arrange(D_term) |>
          mutate(weight = sapply(as.character(D_term), \(x) {
            w <- config$weights$term[[x]]
            if (is.null(w)) "MISSING from config" else as.character(w)
          }))
      }),

    tr_check("2.2", "Records per student, by school",
      good = "Per-school record distributions and expression variety look plausible for HS schedules.",
      evidence = function() {
        d <- cs_data
        if (has_cols("C_course_time_exclude", "C_class_size_and_teacher_load_exclude"))
          d <- d |> filter(C_course_time_exclude == "Include",
                           C_class_size_and_teacher_load_exclude == "Include")
        rps <- d |> count(D_location_name, D_stu_id, name = "records_per_student")
        d |> group_by(D_location_name) |>
          summarise(n_students = n_distinct(D_stu_id),
                    n_teachers = n_distinct(D_employee_id),
                    terms = paste(sort(unique(na.omit(D_term))), collapse = ", "),
                    n_expressions = n_distinct(D_expression, na.rm = TRUE),
                    .groups = "drop") |>
          left_join(rps |> group_by(D_location_name) |>
                      summarise(avg_records_per_student = round(mean(records_per_student), 1), .groups = "drop"),
                    by = "D_location_name") |>
          arrange(D_location_name)
      }),

    tr_check("2.3", "Student demographics — present and populated?",
      good   = "The demographic fields from the 01 merge exist and are well-filled.",
      status = tr_safe(if (all(unlist(config$stu_demographics_join$brings_in) %in% names(cs_data))) "ok" else "none", "none"),
      evidence = function() {
        f <- unlist(config$stu_demographics_join$brings_in)
        present <- f %in% names(cs_data)
        fill <- vapply(f, function(x) if (x %in% names(cs_data)) round(100 * mean(!is.na(cs_data[[x]])), 1) else NA_real_, numeric(1))
        data.frame(field = f, present = ifelse(present, "yes", "no"), pct_filled = fill)
      }),

    tr_check("2.4", "Did the course-coding merge land cleanly?",
      good   = "Unique key (no row multiplication), row count stable pre/post, and every course matched.",
      value  = tr_safe(glue("{merge_diag$n_unmatched} unmatched course rows"), NA),
      status = tr_safe(if (isTRUE(merge_diag$key_unique) && isTRUE(merge_diag$row_stable) && merge_diag$n_unmatched == 0) "ok" else "flag", "none"),
      evidence = function() {
        if (is.null(merge_diag)) return(NULL)
        summ <- data.frame(
          check  = c("key unique (no multiplication)", "rows stable pre/post", "unmatched course rows"),
          result = c(ifelse(merge_diag$key_unique, "yes", "NO"),
                     paste0(format(merge_diag$n_before, big.mark = ","), " -> ",
                            format(merge_diag$n_after,  big.mark = ",")),
                     format(merge_diag$n_unmatched, big.mark = ",")))
        list("Merge summary" = summ,
             "Course names that didn't get coded" = merge_diag$unmatched_courses)
      }),

    tr_check("2.5", "Coding output — sane distributions, nothing left uncoded?",
      good   = glue("Uncoded <= {config$coding$uncoded_flag_threshold_pct}% of records; distributions plausible."),
      value  = tr_safe(paste0(round(100 * mean(cs_data$C_course_subject_area == "UNCODED", na.rm = TRUE), 1), "% uncoded"), NA),
      status = tr_safe(tr_threshold_status(100 * mean(cs_data$C_course_subject_area == "UNCODED", na.rm = TRUE),
                                           config$coding$uncoded_flag_threshold_pct, "below"), "none"),
      evidence = function() {
        sa <- cs_data |> count(C_course_subject_area, name = "n_records") |>
          mutate(pct_records = round(100 * n_records / sum(n_records), 1)) |> arrange(C_course_subject_area)
        sub <- cs_data |> count(C_course_subject_area, C_course_subject, name = "n_records") |>
          mutate(pct_records = round(100 * n_records / sum(n_records), 1)) |> arrange(C_course_subject_area)
        list("Subject Area — % of records" = sa, "Subject — % of records" = sub)
      }),

    tr_check("2.6", "Exclusions — right thing dropping, for the right reason?",
      good = "Every excluded course traces to a coding value that justifies it.",
      evidence = function() {
        cs_data |>
          filter(C_course_time_exclude == "Exclude" | C_class_size_and_teacher_load_exclude == "Exclude") |>
          count(C_course_subject, D_course_name, C_course_time_exclude,
                C_class_size_and_teacher_load_exclude, name = "n_courses", sort = TRUE) |>
          arrange(C_course_subject)
      }),

    tr_check("2.7", "Which expressions carry real instruction?",
      good = "Expressions that are entirely non-standard formats are identified so they fall out.",
      evidence = function() {
        cs_data |>
          count(D_location_name, D_expression, C_course_format, name = "n_records") |>
          pivot_wider(names_from = C_course_format, values_from = n_records, values_fill = 0) |>
          filter(`Standard` == 0) |>
          arrange(D_location_name, D_expression)
      }),

    # ---- double-booking sizing ----
    tr_check("2.8", "Class ID — what defines a class here?",
      good  = "The recipe marks a place-in-time without over-splitting real classes.",
      value = tr_safe(paste(config$class_id$recipe, collapse = " + "), NA),
      evidence = function() {
        cols <- config$class_id$recipe[config$class_id$recipe %in% names(cs_data)]
        cs_data |>
          filter(!is.na(C_class_id)) |>
          group_by(across(all_of(c("C_class_id", cols)))) |>
          summarise(course = dplyr::first(D_course_name),
                    n_students = n_distinct(D_stu_id), .groups = "drop") |>
          arrange(desc(n_students)) |> (\(d) slice_sample(d, n = min(8, nrow(d))))()
      }),

    tr_check("2.9", "Double-booking sizing — how big, and of what kind?",
      good   = "<2-3% may skip; concentrated in a few schools/classes -> hand-resolve; widespread -> full.",
      value  = tr_safe(glue("students {pct_of(length(flagged_students), n_stu_time)}% \u00B7 teachers {pct_of(length(flagged_teachers), n_tch_load)}%"), NA),
      status = tr_safe({
        r <- c(pct_of(length(flagged_students), n_stu_time), pct_of(length(flagged_teachers), n_tch_load))
        if (all(is.na(r))) "none" else if (max(r, na.rm = TRUE) < 3) "ok" else "flag"
      }, "none"),
      evidence = function() {
        incl_s <- cs_long |> filter(C_course_time_exclude == "Include")
        incl_t <- cs_long |> filter(C_class_size_and_teacher_load_exclude == "Include")

        # by-school rates
        stu <- cs_data |> filter(C_course_time_exclude == "Include") |>
          group_by(D_location_name) |>
          summarise(n_students = n_distinct(D_stu_id),
                    n_db = n_distinct(D_stu_id[D_stu_id %in% flagged_students]), .groups = "drop") |>
          mutate(pct_db = pct_of(n_db, n_students)) |> arrange(desc(pct_db))
        tch <- cs_data |> filter(C_class_size_and_teacher_load_exclude == "Include") |>
          group_by(D_location_name) |>
          summarise(n_teachers = n_distinct(D_employee_id),
                    n_db = n_distinct(D_employee_id[D_employee_id %in% flagged_teachers]), .groups = "drop") |>
          mutate(pct_db = pct_of(n_db, n_teachers)) |> arrange(desc(pct_db))

        # sample colliding rows (flat, collisions adjacent, max 25)
        stu_grp <- incl_s |> group_by(D_stu_id, M_conflict_key) |>
          summarise(nc = n_distinct(C_class_id), .groups = "drop") |> filter(nc > 1) |> (\(d) slice_sample(d, n = min(10, nrow(d))))()
        stu_rows <- incl_s |> semi_join(stu_grp, by = c("D_stu_id", "M_conflict_key")) |>
          distinct(D_location_name, D_stu_id, M_conflict_key, C_class_id, D_term, D_expression, D_course_section_id, D_course_name) |>
          arrange(D_location_name, D_stu_id, M_conflict_key) |>
          transmute(D_location_name, D_stu_id, D_term, D_expression, D_course_section_id, D_course_name, M_conflict_key)
        # Teacher collisions with the largest combined student rosters
        tch_collision_rows <- incl_t |>
          distinct(
            D_location_name, D_employee_id, M_conflict_key,
            C_class_id, D_stu_id, D_term, D_expression,
            D_course_section_id, D_course_name
          )

        tch_grp <- tch_collision_rows |>
          group_by(D_location_name, D_employee_id, M_conflict_key) |>
          summarise(
            n_classes = n_distinct(C_class_id),
            n_student_rows = n(),
            n_unique_students = n_distinct(D_stu_id),
            .groups = "drop"
          ) |>
          filter(n_classes > 1) |>
          slice_max(n_student_rows, n = 10, with_ties = FALSE)

        tch_rows <- tch_collision_rows |>
          semi_join(
            tch_grp,
            by = c("D_location_name", "D_employee_id", "M_conflict_key")
          ) |>
          group_by(D_location_name, D_employee_id, M_conflict_key, C_class_id) |>
          summarise(
            D_term             = paste(sort(unique(na.omit(D_term))), collapse = ", "),
            D_expression       = paste(sort(unique(na.omit(D_expression))), collapse = ", "),
            D_course_section_id = paste(sort(unique(na.omit(D_course_section_id))), collapse = ", "),
            D_course_name      = paste(sort(unique(na.omit(D_course_name))), collapse = " | "),
            n_students_in_class = n_distinct(D_stu_id),
            .groups = "drop"
          ) |>
          left_join(
            tch_grp |> select(D_location_name, D_employee_id, M_conflict_key, n_student_rows, n_unique_students),
            by = c("D_location_name", "D_employee_id", "M_conflict_key")
          ) |>
          arrange(desc(n_student_rows), D_location_name, D_employee_id, M_conflict_key, desc(n_students_in_class)) |>
          select(D_location_name, D_employee_id, D_term, D_expression, D_course_section_id, D_course_name,
                 M_conflict_key, n_students_in_class, n_student_rows, n_unique_students)

        # class-level 100% + the class(es) their students collide with
        cls <- cs_data |> filter(C_course_time_exclude == "Include") |>
          distinct(C_class_id, D_course_name, D_location_name, D_stu_id) |>
          group_by(C_class_id, D_course_name, D_location_name) |>
          summarise(n_students = n_distinct(D_stu_id),
                    n_db = n_distinct(D_stu_id[D_stu_id %in% flagged_students]), .groups = "drop") |>
          mutate(pct_db = pct_of(n_db, n_students)) |>
          filter(pct_db == 100) |> arrange(desc(n_students)) |> head(20)
        kc  <- incl_s |> distinct(D_stu_id, M_conflict_key, C_class_id)
        tgt <- kc |> filter(C_class_id %in% cls$C_class_id) |> distinct(D_stu_id, M_conflict_key)
        at  <- kc |> semi_join(tgt, by = c("D_stu_id", "M_conflict_key"))
        look <- cs_data |> distinct(C_class_id, D_course_name)
        partners <- at |> filter(C_class_id %in% cls$C_class_id) |> rename(focus = C_class_id) |>
          inner_join(rename(at, partner = C_class_id), by = c("D_stu_id", "M_conflict_key"), relationship = "many-to-many") |>
          filter(focus != partner) |>
          left_join(rename(look, partner = C_class_id, partner_course = D_course_name), by = "partner") |>
          distinct(focus, M_conflict_key, partner, partner_course) |>
          group_by(focus) |>
          summarise(collides_with = paste(head(unique(paste0(partner_course, " @", M_conflict_key)), 3), collapse = "; "),
                    .groups = "drop")
        cls <- cls |> left_join(rename(partners, C_class_id = focus), by = "C_class_id")

        cls_db <- cs_data |> filter(C_course_time_exclude == "Include") |>
          group_by(D_location_name) |>
          summarise(n_classes = n_distinct(C_class_id),
                    n_db = n_distinct(C_class_id[C_class_id %in% cls$C_class_id]), .groups = "drop") |>
          mutate(pct_db = pct_of(n_db, n_classes)) |> arrange(desc(pct_db))

        list("Sizing filters applied" = sizing_filter_summary(as.data.frame(cs_data), config$db_sizing_filters),
             "Student double-bookings by school" = stu,
             "Student collisions \u2014 sample rows (max 20)" = stu_rows,
             "Teacher double-bookings by school" = tch,
             "Teacher collisions \u2014 sample rows (max 20)" = tch_rows,
             "Class double-bookings by school" = cls_db,
             "Class-level (100%) double-bookings (max 20)" = cls)
      })
  )
)

htmltools::browsable(render_stage(s02))
```

*Next -> stage 03: TPR weights, class weight, and double-booking resolution.*

## Re-export for the next stage

```{r export}
#| eval: false
ers_write_sharepoint(
  data = cs_data,
  folder_path = raw_data_folder_path,
  file_name_with_extension = "/2. Processed Data/02_course_data_post_code_and_exclude.csv")
```
