---
title: "Course Schedule Processing — Methodology Review"
subtitle: "Springfield · 2024–25"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(gt)
library(dplyr)
library(yaml)
library(erstools)          # ers_read_sharepoint()
source("tier_review.R")
config <- yaml::read_yaml("district_config.yml")

# --- SharePoint connection (mirrors script 01) ---
orgfiles_site_url <- "https://erstrategies1.sharepoint.com/sites/orgfiles"
orgfiles_site     <- get_sharepoint_site(site_url = orgfiles_site_url)
internal_drive    <- orgfiles_site$get_drive("Internal")

# --- Folder path + raw (pre-explosion) loads ---
raw_data_folder_path <- "District Partners/Springfield MA/Springfield - 26 - HS Scheduling/1. Data & Analysis - Secure/For R"

cs_raw_file <- ers_read_sharepoint(
  folder_path = raw_data_folder_path,
  file_name_with_extension = "/Raw Data/student_course_schedules.csv",  # <- your exact pre-explosion file
  drive_name = "internal_drive")

stu_demographics <- ers_read_sharepoint(
  folder_path = raw_data_folder_path,
  file_name_with_extension = "/Raw Data/student_enrollment_and_demographic.xlsx",
  drive_name = "internal_drive")

# --- Map district headers -> ERS D_ names (single source: config$column_map) ---
# rename(any_of()) renames only columns that are present, so a missing/renamed
# district column is skipped rather than erroring (non-blocking, by design).
cs_raw_file      <- dplyr::rename(cs_raw_file,      dplyr::any_of(unlist(config$column_map$course_schedule)))
stu_demographics <- dplyr::rename(stu_demographics, dplyr::any_of(unlist(config$column_map$demographics)))
```

## How to read this

Each tier answers one readiness question. Every entry is a **decision** — a
yardstick plus the recorded call — and where it helps, an **evidence table**
sits directly beneath so you can gut-check the data behind the decision.
Statuses never stop the render: a missing input (or unloaded data) shows a flag,
not a crash. An amber **note** is an analyst override. And the golden rule:
**processing computes flags, it never deletes** — every exclusion is a column a
later metric can honor or ignore.

## Tier 0 — Raw CS data scan

*Did we get the data we asked for, and what shape is it in?*

```{r tier0}
t0 <- tr_tier(
  number = 0,
  title  = "Raw CS data scan",
  question = "Did we get the data we asked for, and what shape is it in?",
  checks = list(

    tr_check("0.1", "Snapshot or cumulative enrollment?",
      good    = "Unambiguous; if cumulative, a snapshot date is agreed with the district.",
      value   = config$intake$enrollment_type,
      status  = "ok",
      records = "Snapshot as-received"),

    tr_check("0.2", "Do all critical CS fields exist and map to ERS D_ names?",
      good    = "No critical field MISSING after mapping.",
      value   = tr_safe(cs_raw_file |> select_if(~ any(. == "MISSING")) |> ncol(), NA),
      status  = "todo",   # then: tr_threshold_status(value, 0, "below")
      records = "List any missing criticals -> data-request follow-ups"),

    tr_check("0.3", "Expression, or period + rotation separately?",
      good    = "We know which; an expression (e.g. 2-8(A)) is flagged to parse in Tier 1.",
      value   = config$intake$time_field,
      status  = "ok",
      records = "Expression -> parse in Tier 1",
      evidence = function() {
        cs_raw_file |>
          count(D_expression, name = "n_records", sort = TRUE)
      }),

    tr_check("0.4", "Term structure — full-year, semesterized, or a mix?",
      good    = "Known; a mix is expected and noted (drives overlap-aware DB sizing).",
      value   = "FY + semester (mix)",
      status  = "flag",
      records = "Mixed FY/semester -> split tables needed in Tier 2",
      note    = "Mix is expected — flagged only so Tier 1 DB sizing runs interval-overlap-aware, not string-equality.",
      evidence = function() {
        configured <- names(config$weights$term)
        cs_raw_file |>
          count(D_term, name = "n_records", sort = TRUE) |>
          mutate(in_config = if_else(D_term %in% configured, "yes", "MISSING from config"))
      }),

    tr_check("0.5", "Records per student vs. bell schedule?",
      good    = "Per-school counts and pre-explosion ranges reconcile with the bell schedule.",
      value   = tr_safe(
                  cs_raw_file |> count(D_stu_id) |> count(n) |>
                    slice_max(nn, n = 1, with_ties = FALSE) |> pull(n), NA),
      status  = "todo",
      records = "Modal records/student per school + reconciliation note against bell schedule",
      evidence = function() {
        modal <- cs_raw_file |>
          count(D_location_name, D_stu_id, name = "recs") |>
          count(D_location_name, recs, name = "n") |>
          group_by(D_location_name) |>
          slice_max(n, n = 1, with_ties = FALSE) |>
          transmute(D_location_name, modal_records = recs)

        cs_raw_file |>
          group_by(D_location_name) |>
          summarise(
            n_students  = n_distinct(D_stu_id),
            n_teachers  = n_distinct(D_employee_id),
            terms       = paste(sort(unique(D_term)), collapse = ", "),
            n_rotations = n_distinct(D_expression),
            n_periods   = n_distinct(D_period),
            .groups = "drop") |>
          left_join(modal, by = "D_location_name") |>
          arrange(D_location_name)
      }),

    tr_check("0.6", "Data inventory — student demographics (grade, SWD/ELL/FRL)",
      good    = "Present; unique-student counts reconcile with the CS file per school.",
      value   = "present · separate file",
      status  = tr_presence_status(present = TRUE),
      records = "Merge + student-drop decision happen in Tier 1 step 0",
      evidence = function() {
        loc <- cs_raw_file |> distinct(D_location_id, D_location_name)

        compare <- loc |>
          left_join(stu_demographics |> distinct(D_stu_id, D_location_id) |>
                      count(D_location_id, name = "n_students_demo"),
                    by = "D_location_id") |>
          left_join(cs_raw_file |> distinct(D_stu_id, D_location_id) |>
                      count(D_location_id, name = "n_students_cs"),
                    by = "D_location_id") |>
          mutate(diff_cs_minus_demo = n_students_cs - n_students_demo) |>
          arrange(D_location_name)

        rates <- stu_demographics |>
          summarise(
            n_students = n_distinct(D_stu_id),
            pct_swd = tr_pct_true(D_stu_swd_flag),
            pct_ell = tr_pct_true(D_stu_ell_flag),
            pct_frl = tr_pct_true(D_stu_poverty_flag))

        list(
          "Unique students per school — demographics vs CS file" = compare,
          "Demographic rates (demographics file only, pre-merge)" = rates
        )
      }),

    tr_check("0.7", "Data inventory — proficiency source (student grades/scores)",
      good    = "Present for the same students/years.",
      value   = "not yet received",
      status  = tr_presence_status(present = FALSE),
      records = "Follow up with district — needed for Tier 3 proficiency rollup",
      note    = "Override: running the schedule side now; proficiency joins at Tier 3 without re-running Tiers 0-2."),

    tr_check("0.8", "Data inventory — teacher experience (licensure/HR)",
      good    = "Present and joinable on teacher ID.",
      value   = "partial",
      status  = tr_presence_status(present = TRUE, partial = TRUE),
      records = "Have licensure; HR FTE outstanding -> follow up"),

    tr_check("0.9", "Double bookings are the gate ahead (not sized here)",
      good    = "Signpost only — sizing needs Tier 1 coding + exclusions + relevant periods.",
      status  = "na",
      records = "DB sized at end of Tier 1, once we know which periods count")
  )
)

render_tier(t0)
```

## Tier 1 — Student & course coding

*What is each course & student, what's excluded, and how real are the DBs?*

```{r tier1}
t1 <- tr_tier(
  number = 1,
  title  = "Student & course coding",
  question = "What is each course & student, what's excluded, and how real are the DBs?",
  checks = list(

    tr_check("1.0", "Merge demographics & set the student-drop decision",
      good    = "1 row per student in the demo file; every course record matches, or non-matches understood.",
      status  = "todo",
      records = "Merge state + who is dropped for missing demographics & how many (born ONCE here)"),

    tr_check("1.1", "Code the courses (Subject, Rigor, Time of Day, Class Format, metric booleans)",
      good    = glue::glue("Uncoded <= {config$coding$uncoded_flag_threshold_pct}% of records — intentional residuals overridden with a note."),
      status  = "todo",
      records = "Coding decisions + any course resolved using demographics (the 'Essentials = %SWD/ELL' test)"),

    tr_check("1.2", "Derive exclusions from coding (Time / Class Size / Teacher Load)",
      good    = "Every rule intentional; exclusions are flags, not deletions.",
      status  = "todo",
      records = "Exclusion rules applied + counts of rows FALSE per metric"),

    tr_check("1.3", "Identify which periods count",
      good    = "Every non-instructional period deliberately excluded (the Sci-Tech judgment).",
      value   = paste(config$relevant_periods$include, collapse = ", "),
      status  = "todo",
      records = "Periods in/out + why (carries forward the Tier 0 candidates)",
      evidence = function() {
        # courses per period supports the include/exclude call:
        # low-unique-course periods are the ones to scrutinize (advisory/afterschool/counselor).
        cs_raw_file |>
          group_by(D_period) |>
          summarise(n_unique_courses = n_distinct(D_course_name),
                    n_records = n(), .groups = "drop") |>
          arrange(n_unique_courses)
      }),

    tr_check("1.4", "Size all three DB types (student / teacher / class)",
      good    = "Overlap-aware: term = interval intersection, rotation = set intersection. <2–3% may skip; concentrated -> hand; widespread -> full.",
      status  = "todo",
      records = "DB size by type + concentration + provisional PER-TYPE resolve decision (skip / hand / full)")
  )
)

render_tier(t1)
```

---

*Evidence attaches to any check the same way — pass `evidence = function() {...}`
returning a data frame (or a named list of frames). Tiers 2–3 stamp out through
the same `render_tier()`.*
